Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-21 ThesisJP/Background/background.tex
Stil writing...

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{img/free_lunsh.png}
\caption{Illustration of transitions between different eras in the computer industry. Currently there are three ongoing transitions; multi-core, heterogeneous-core and cloud-core. Image courtesy of Herb Sutter (herbsutter.com).}
\label{fig:jungle}
\end{figure}

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) instruction \cite{Flynn1966}, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD Extensions}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced Vector Extensions}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers which means that eight 32 bit floating point values can be processed per clock cycle. This extension is found in the latest CPUs from Intel. AVX-512 extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs. The latest quad-core CPU from Intel, the Haswell architecture, can therefore process 32 32 bit floating point values per clock cycle. With a clock frequency of 3.9 GHz this sums up to a theoretical peak single precision performance of 124.8 giga single precision floating point operations per second (GFLOPS). In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing 16 32 bit floating point values at 1.1 GHz. The total theoretical peak single precision FLOPS for a high-end Haswell CPU is therfore $124.8 + (40*16*1.1) = 828.8$ GFLOPS as visualized in Fig\,\ref{fig:cpu_vs_gpu}.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors. The Fermi GPU architecture by Nvidia, which is used the most in this theses, have up to 16 SIMD units capable of performing 32 32 bit floats per clock cycle at 772 MHz. The theoretical peak single precision FLOPS for this architecture is by that found to be $16*32*772 = $ (needs to add FMA and SP-units, What about FMA on the CPU?)

It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.

The SIMD registers on a GPU is longer than on a CPU and the GPU have more of them.

 the most powerful integrated intel GPU (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).

About theoretical FLOPS: Note that these are theoretical numbers and the actual throughput typically is algorithm dependent.

Compare CPU and GPU ALUs. SSE, AVX ... The GPU has super-wide registers. The SM in Fermi have 32 cores or 1024 bit registers. The SMX in Kepler have 192 cores. 6144 bit registers. SSE (Streaming SIMD Extensions) instructions have 128 bit registers. AVX (Advanced Vector Extensions) have 256 bit registers. AVX-512 has 512 bit registers. Single instruction, multiple data (SIMD).

In this section we have focused on CPUs from Intel and GPUs from Nvidia, note that GPUs and CPUs from AMD have similar features.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{img/CPU-GPU.png}
\caption{Overview of CPU and GPU architecture.}
\label{fig:cpu_gpu}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/free_lunsh.png}
\caption{Illustration of transitions between different eras in the computer industry. Currently there are three ongoing transitions; multi-core, heterogeneous-core and cloud-core. Image courtesy of Herb Sutter (herbsutter.com).}
\label{fig:jungle}
\end{figure}

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction \cite{Flynn1966}, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{img/CPU-GPU.png}
\caption{Overview of CPU and GPU architecture. (make my own)}
\label{fig:cpu_gpu}
\end{figure}

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four 32 bit floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instructions sets have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs. The latest CPU architecture from Intel, the Haswell architecture, has two FMA ports and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak single precision performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture. In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz, 
\begin{equation}
16*40*1.1 = 704\,\text{GFLOPS}.
\end{equation}

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The latest GPU architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. This GPU is also capable of performing FMA instructions, doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (cos, exp, ect.)}) which add close to 1 TFLOPS in additional power. We see how the wide SIMD register found on GPUs results in five times the performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that actual throughput typically is algorithm dependent. In this section we have focused on CPUs from Intel and GPUs from Nvidia, however, GPUs and CPUs from AMD have similar specifications.

@@

### 2014-02-21 ThesisJP/Introduction/introduction.tex
Stil writing...

\subsection{Paper\,I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, submitted.}\\\\
The first paper describes in details how the Capon beamformer was mapped to GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements. This makes it somewhat easier to reach the goal of real-time processing.

@@

\subsection{Paper\,III}
\textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
\textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted.}\\\\
In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer. As mentioned, shift-invariant imaging is essential if the method is ever to be adopted. The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations. The point scatterers were observed to twinkle like stars in the sky. In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without being clearly stated. Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects. Smooth presentation of probe motion is equally important since real-time user interaction is one of the key selling points of ultrasound imaging. The frame rate should therefore be higher than 10 frames per second.

>>>>

\subsection{Paper\,I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
The first paper describes in details how the Capon beamformer was mapped to GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements. This makes it somewhat easier to reach the goal of real-time processing.

@@

\subsection{Paper\,III}
\textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
\textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication.}\\\\
In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer. As mentioned, shift-invariant imaging is essential if the method is ever to be adopted. The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations. The point scatterers were observed to twinkle like stars in the sky. In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without being clearly stated. Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects. Smooth presentation of probe motion is equally important since real-time user interaction is one of the key selling points of ultrasound imaging. The frame rate should therefore be higher than 10 frames per second.

### 2014-02-21 ThesisJP/Thesis.tex
Continued writing background information.
writing and added figure.
Stil writing...

@@

This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
\begin{enumerate}[I]
\renewcommand\labelenumi{\bfseries\theenumi}
	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, submitted}.
	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}.
	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
\end{enumerate}

@@

%%%% Paper 1 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
\authors{
	Jo Inge Buskenes$^{1}$, 
    	\textbf{Jon Petter \AA{}sen}$^{2}$, 
    	Carl-Inge C. Nilsen$^{1}$ and
    	Andreas Austeng$^{1}$
} 
{
	$^{1}$Department of Informatics, University of Oslo, Oslo, Norway\\
    	$^{2}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway
}
\noindent \textit{IEEE Transactions on Oceanic Engineering, submitted}.
%\newpage\input{./CaponSonarGPU/CaponSonarGPU}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./CaponSonarGPU/bibCaponSonarGPU]
\end{bibunit}

@@

%%%% Paper 3 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
\runningtitle{Capon Beamforming and Moving Objects}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$, 
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
%\input{./CaponMovingObjects/CaponMovingObjects}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./CaponMovingObjects/bibCaponMovingObjects]
\end{bibunit}

>>>>

% custom commands
\newcommand{\mat}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\degree}{\ensuremath{^\circ}}
\newcommand{\R}{$\mat{\hat{R}}$ }

@@

This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
\begin{enumerate}[I]
\renewcommand\labelenumi{\bfseries\theenumi}
	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, accepted for publication}.
	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008}.
	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
\end{enumerate}

@@

%%%% Paper 1 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
\authors{
	Jo Inge Buskenes$^{1}$, 
    	\textbf{Jon Petter \AA{}sen}$^{2}$, 
    	Carl-Inge C. Nilsen$^{1}$ and
    	Andreas Austeng$^{1}$
} 
{
	$^{1}$Department of Informatics, University of Oslo, Oslo, Norway\\
    	$^{2}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway
}
\noindent \textit{IEEE Transactions on Oceanic Engineering, accepted for publication}.
%\newpage\input{./CaponSonarGPU/CaponSonarGPU}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./CaponSonarGPU/bibCaponSonarGPU]
\end{bibunit}

@@

%%%% Paper 3 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
\runningtitle{Capon Beamforming and Moving Objects}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$, 
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication}.
%\input{./CaponMovingObjects/CaponMovingObjects}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./CaponMovingObjects/bibCaponMovingObjects]
\end{bibunit}
